Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-11-01 Writing/manuscript.Rnw
manuscript.rnw created

@@

@@

Figure \ref{fig:yield-dif} shows the difference in yield of the second and third (highest) aquifer quantiles relative to the first (lowest) quantile with the 95\% confidence interval for corn and soybean. For corn, in years with water deficit greater than about 700 mm, the highest aquifer thickness group has statistically significantly higher yields than the lowest aquifer thickness group. At an extreme water deficit level of 950 mm (average water deficit in 2012, the most extreme drought in our historical record), the difference in yields between the smallest and largest aquifer thickness groups is 1.15 tonne/ha. Unlike corn, irrigated soybean yields are not statistically distinguishable among the three aquifer thickness groups. This is likely because soybean is not as water demanding a crop as corn, meaning that even counties with lower aquifer thickness can meet the majority of crop water demands as well yield is not a significantly binding constraint on irrigation management decisions.

% At low water deficits (below 200mm), Figure \ref{fig:yield-dif} illustrates that counties with the highest levels of aquifer thickness (i.e., 3rd quantile in Figure \ref{fig:yield-response}) experience statistically lower yields for corn and soybean than counties with the lowest level of aquifer thickness (1st quantile). In case of soybean, irrigated yields for the lowest aquifer thickness category converge to levels comparable to those obtained from rainfed production. These counter-intuitive results reflect a combination of two factors. First, in years with small water deficits there is less need for irrigation and hence irrigated and rainfed yield differences are reduced. Second, years with small or even negative water deficits are associated with wetter conditions during the growing season, which can negatively impact yield if waterlogging occurs leading to greater risks of physical crop damage (e.g., lodging of grains), delays to planting/harvesting, nutrient leaching, and restricted root growth in waterlogged soils \citep{li2019excessive}. Declines in irrigated yields could be indicative of over-irrigation by farmers, which has been shown to occur in the HPA region in wet years \citep{foster2019assessing, gibson2017case, gibson2019benchmarking}, exacerbating these risks. However, the small declines in irrigated yields at low or negative levels of water deficit may also reflect the fact that wetter years are typically cooler, reducing positive heat effects on crop growth and yield formation.

@@

{\centering \includegraphics[width=6.5in,height=500px,]{../../Figures/g_yield_intensive}

@@

\caption{The impact of water deficit and aquifer thickness on rainfed and irrigated per-area yields of corn and soybean in US High Plains. Irrigated yield responses to water deficits are dissagregated by aquifer thickness level, considering three quantile groups that capture the range of aquifer thickness values in our sample: (i) 9 m to 30.1 m; (ii) 30.1 m to 67.2 m; and (iii) 67.2 m to 215.8 m. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop. Total number of observations are $8,773$ and $5,977$ for corn and soybean regressions, respectively.}\label{fig:yield-response}
\end{figure}

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_ir_yield_dif}

@@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around 0.90 (90\%) for an average county-level aquifer thickness of 100 m to 0.72 (72\%) for an average county-level aquifer thickness of 10 m (See Appendix \ref{test-dif-share} for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean.

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_share}

@@

\caption{The estimated impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean producing counties in the US High Plains (Note: shaded area represents 95\% confidence interval). The histograms on the top of each panel show the distribution of county-level aquifer thickness values included for each crop. Total number of observations are $2,523$ and $1,558$ for corn and soybean regressions, respectively.}\label{fig:ir-share}
\end{figure}

@@

@@

Figure \ref{fig:tot-impact} and Figure \ref{fig:dif-tot-impact} also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950 mm, for example, reducing from 100 m aquifer thickness to 70 m (decline of 30 m) would result in only 0.09 tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \ref{fig:ir-share}. Reducing from 70 m aquifer thickness to 40 m (decline of 30 m) would result in 0.95 tonne/ha of production loss for corn, with a further drop from 40 m aquifer thickness to 10 m (decline of 30 m) leading to 1.57 tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \ref{fig:ir-share}), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers' ability to effectively and reliably buffer crops against water deficits during the growing season.

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_total_impact}

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_avg_yield_dif}

@@

{\centering \includegraphics[width=750px,height=6in,]{../../Figures/g_y_wd_all}

@@

As a proxy for the extent of aquifer depletion we use annual estimates of aquifer thickness generated by existing research \citep{haacker2016water,haacker2023}. These data provide annual estimates of the aquifer thickness at 250m resolution since 1935. Annual aquifer thickness maps were derived based on interpolation of 14,000 to 20,000 individual observations of water levels per year, along with detailed maps of both the land surface elevation and bedrock elevation of the HPA. Aquifer thickness estimates for each year were aggregated to the county level using an area-weighted average to match the resolution of agricultural production data. Aquifer thickness values were only estimated for counties with irrigated yield observations. Furthermore, during the aggregation process, any aquifer cells within a county that had a aquifer thickness value lower than 9 m were removed before aggregation to reflect the fact that aquifer thicknesses lower than this threshold are not considered to be viable for extraction and therefore it is unlikely these areas contribute groundwater for irrigation production \citep{fenichel2016measuring, haacker2016water, deines2020transitions}. Figure \ref{fig:sat-map} presents the distribution of aquifer thickness across our study area for counties included in corn and soybean analyses. Aquifer thickness levels vary substantially over space and time, which we will exploit in our analysis to quantify how aquifer thickness influences farmers' exposure and responses to drought.

These data were combined to form four sets of data for regression analysis: (1) per-acre corn yield ($8,773$ observations), (2) per-acre soybean yield ($5,977$ observations), (3) share of irrigated corn ($2,523$ observations), and (4) share of irrigated soybean ($1,558$ observations).

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_map}

@@

\hypertarget{data-and-code-availability}{%
\subsection{Data Availability}\label{data-and-code-availability}}

All the codes used for this study are available at the following Github repository (\href{https://github.com/tmieno2/Drought-Production-Risk-Aquifer}{link}). All analyses were performed in R \citep{R} and all the raw data-sets used in this study are publicly accessible. The boundary shape file of the High Plains aquifer was obtained from the U.S. Geological Survey at \url{https://water.usgs.gov/GIS/metadata/usgswrd/XML/ds543.xml#stdorder}. This data is used for Figure \ref{fig:sat-map}. Corn and soybean yield data were obtained from the U.S. Department of Agriculture's National Agricultural Statistics Service (USDA-NASS), available at \url{https://www.nass.usda.gov/Quick_Stats/}. The R tidyUSDA package \citep{RtidyUSDA} was used to automatically download the relevant data. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. Weather data (precipitation, temperature, and evapotranspiration) was obtained from the gridMET dataset \citep{Abatzoglou2013}, which is available at \url{https://www.climatologylab.org/gridmet.html}. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. The aquifer thickness data were obtained from Hydroshare \citep{haacker2023} available at \url{https://www.hydroshare.org/resource/7d925c7944244032af98c9ed20c22db6/}. This data is used for Figure \ref{fig:sat-map} and the regression analysis. Soil characteristics data was obtained from the Soil Survey Geographic Dataset, which is used for regression analysis. The R soilDB package \citep{Rsoildb} was used to download the data. Regression analyses are implemented using the R fixest package \citep{Rfixest}. Other key R packages include ggplot2 \citep{Rggplot2} for creating plots, data.table \citep{Rdatatable} for data wrangling, and sf \citep{Rsf} for spatial data operations. The data used for the regression analysis can be generated using our codes by following the instructions available at the linked Github repository.

\subsection{Code Availability}

@@

\section*{Author Contribution Statement}

\section*{Competing Interests Statement}

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_yield_with_conf_corn}

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_yield_with_conf_soy}

@@

{\centering \includegraphics[width=6.5in,height=500px,]{../../Figures/g_avg_yield}

@@

Figure \ref{fig:test-sat-share} presents the difference in the estimated share of irrigated acres at different levels of aquifer thickness relative to the aquifer thickness level of 100 m, which represents a high level of aquifer thickness for our study area that exerts no constraints on water supply for irrigation. Grey shaded areas represented the 95\% confidence intervals for the estimated relationships. Where this range exceeds the red zero line, this indicates that the difference in aquifer thickness (relative to 100 m baseline) leads to a statistically significant (at 5\% level) reduction in irrigated production area share.

@@

{\centering \includegraphics[width=6.5in,height=500px,]{../../Figures/g_share_dif}

@@

\caption{Decline in the share of acres that are irrigated as a function of different levels of aquifer thickness relative to a baseline aquifer thickness level of 100 m (black line). The grey shaded areas represent the 95\% confidence interval for this estimated relationship.}\label{fig:test-sat-share}
\end{figure}

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_variation}

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_share_comp}

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_basis}

@@

{\centering \includegraphics[width=6in,height=500px,]{../../Figures/g_il_y}

>>>>

<<init, echo = FALSE>>=
knitr::opts_chunk$set(
  echo = FALSE
)
library(tidyverse)
library(data.table)
@

@@

<<load_results, echo = FALSE>>=
setwd(here::here("Writing"))
all_results <- readRDS("../Results/all_results.rds")
@

@@

<<echo = FALSE>>=
delta_yield_lowest_highest <- 
  all_results[crop == "corn", yield_dif_test][[1]] %>%
  .[comp_q == 3, ] %>%  
  .[which.min(abs(balance -950)), dif_hat] %>%
  round(digits = 2)
@

Figure \ref{fig:yield-dif} shows the difference in yield of the second and third (highest) aquifer quantiles relative to the first (lowest) quantile with the 95\% confidence interval for corn and soybean. For corn, in years with water deficit greater than about 700 mm, the highest aquifer thickness group has statistically significantly higher yields than the lowest aquifer thickness group. At an extreme water deficit level of 950 mm (average water deficit in 2012, the most extreme drought in our historical record), the difference in yields between the smallest and largest aquifer thickness groups is \Sexpr{delta_yield_lowest_highest} tonne/ha. Unlike corn, irrigated soybean yields are not statistically distinguishable among the three aquifer thickness groups. This is likely because soybean is not as water demanding a crop as corn, meaning that even counties with lower aquifer thickness can meet the majority of crop water demands as well yield is not a significantly binding constraint on irrigation management decisions.

@@

{\centering \includegraphics[width=6.5in,height=500px,]{../Results/Figures/g_yield_intensive}

@@

<<num_obs_yield_reg, echo = FALSE>>=
nobs_corn <- 
  all_results[crop == "corn", data][[1]] %>% 
  nrow() %>%
  format(big.mark=",")

nobs_soy <- all_results[crop == "soy", data][[1]] %>% 
  nrow() %>%
  format(big.mark=",")
@

\caption{The impact of water deficit and aquifer thickness on rainfed and irrigated per-area yields of corn and soybean in US High Plains. Irrigated yield responses to water deficits are dissagregated by aquifer thickness level, considering three quantile groups that capture the range of aquifer thickness values in our sample: (i) 9 m to 30.1 m; (ii) 30.1 m to 67.2 m; and (iii) 67.2 m to 215.8 m. The histograms on the top of each panel show the distribution of county-level annual water deficit values included in regressions for each crop. Total number of observations are $\Sexpr{nobs_corn}$ and $\Sexpr{nobs_soy}$ for corn and soybean regressions, respectively.}\label{fig:yield-response}
\end{figure}

@@

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_ir_yield_dif}

@@

<<ir_share, echo = FALSE>>=
corn_share <- all_results[crop == "corn", share_pred_data][[1]]

share_high <-
  corn_share[which.min(abs(sat - 130)), ir_share_hat] %>%
  round(digits = 2)

share_low <-
  corn_share[which.min(abs(sat - 10)), ir_share_hat] %>%
  round(digits = 2) 
@

Figure \ref{fig:ir-share} shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \ref{fig:ir-share} demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around \Sexpr{share_high} (\Sexpr{share_high * 100}\%) for an average county-level aquifer thickness of 100 m to \Sexpr{share_low} (\Sexpr{share_low * 100}\%) for an average county-level aquifer thickness of 10 m (See Appendix \ref{test-dif-share} for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean.

@@

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_share}

@@

<<num_obs_share_reg, echo = FALSE>>=
nobs_corn_share <- 
  all_results[crop == "corn", share_reg_data][[1]] %>% 
  nrow() %>%
  format(big.mark=",")

nobs_soy_share <- 
  all_results[crop == "soy", share_reg_data][[1]] %>% 
  nrow() %>%
  format(big.mark=",")
@

\caption{The estimated impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean producing counties in the US High Plains (Note: shaded area represents 95\% confidence interval). The histograms on the top of each panel show the distribution of county-level aquifer thickness values included for each crop. Total number of observations are $\Sexpr{nobs_corn_share}$ and $\Sexpr{nobs_soy_share}$ for corn and soybean regressions, respectively.}\label{fig:ir-share}
\end{figure}

@@

<<res_at_950, echo = FALSE>>=
res_at_950 <-
  all_results[, .(crop, avg_yield_pred_data)] %>%
  unnest(cols = c(avg_yield_pred_data)) %>%
  data.table() %>%
  .[crop == "corn", ] %>%
  .[(abs(balance - 950)) < 5, ]
@

@@

Figure \ref{fig:tot-impact} and Figure \ref{fig:dif-tot-impact} also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950 mm, for example, reducing from 100 m aquifer thickness to 70 m (decline of 30 m) would result in only \Sexpr{round(res_at_950[sat == 100 & crop == "corn", avg_yield] - res_at_950[sat == 70 & crop == "corn", avg_yield], digits = 2)} tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \ref{fig:ir-share}. Reducing from 70 m aquifer thickness to 40 m (decline of 30 m) would result in \Sexpr{round(res_at_950[sat == 70 & crop == "corn", avg_yield] - res_at_950[sat == 40 & crop == "corn", avg_yield], digits = 2)} tonne/ha of production loss for corn, with a further drop from 40 m aquifer thickness to 10 m (decline of 30 m) leading to \Sexpr{round(res_at_950[sat == 40 & crop == "corn", avg_yield] - res_at_950[sat == 10 & crop == "corn", avg_yield], digits = 2)} tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \ref{fig:ir-share}), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers' ability to effectively and reliably buffer crops against water deficits during the growing season.

@@

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_total_impact}

@@

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_avg_yield_dif}

@@

{\centering \includegraphics[width=750px,height=6in,]{../Results/Figures/g_y_wd_all}

@@

As a proxy for the extent of aquifer depletion we use annual estimates of aquifer thickness generated by existing research \citep{haacker2016water,haacker2023}. These data provide annual estimates of the aquifer thickness at 250m resolution since 1935. Annual aquifer thickness maps were derived based on interpolation of 14,000 to 20,000 individual observations of water levels per year, along with detailed maps of both the land surface elevation and bedrock elevation of the HPA. Aquifer thickness estimates for each year were aggregated to the county level using an area-weighted average to match the resolution of agricultural production data. Aquifer thickness values were only estimated for counties with irrigated yield observations. Furthermore, during the aggregation process, any county that had an aquifer thickness value lower than 9 m was removed before aggregation to reflect the fact that aquifer thicknesses lower than this threshold are not considered to be viable for extraction and therefore it is unlikely these areas contribute groundwater for irrigation production \citep{fenichel2016measuring, haacker2016water, deines2020transitions}. Figure \ref{fig:sat-map} presents the distribution of aquifer thickness across our study area for counties included in corn and soybean analyses. Aquifer thickness levels vary substantially over space and time, which we will exploit in our analysis to quantify how aquifer thickness influences farmers' exposure and responses to drought.

These data were combined to form four sets of data for regression analysis: (1) per-acre corn yield ($\Sexpr{nobs_corn}$ observations), (2) per-acre soybean yield ($\Sexpr{nobs_soy}$ observations), (3) share of irrigated corn ($\Sexpr{nobs_corn_share}$ observations), and (4) share of irrigated soybean ($\Sexpr{nobs_soy_share}$ observations).

@@

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_map}

@@

\hypertarget{data-availability}{%
\subsection{Data Availability}\label{data-availability}}

All the raw data-sets used in this study are publicly accessible. The boundary shape file of the High Plains aquifer was obtained from the U.S. Geological Survey at \url{https://water.usgs.gov/GIS/metadata/usgswrd/XML/ds543.xml#stdorder}. This data is used for Figure \ref{fig:sat-map}. Corn and soybean yield data were obtained from the U.S. Department of Agriculture's National Agricultural Statistics Service (USDA-NASS), available at \url{https://www.nass.usda.gov/Quick_Stats/}. The R tidyUSDA package \citep{RtidyUSDA} was used to automatically download the relevant data. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. Weather data (precipitation, temperature, and evapotranspiration) was obtained from the gridMET dataset \citep{Abatzoglou2013}, which is available at \url{https://www.climatologylab.org/gridmet.html}. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. The aquifer thickness data were obtained from Hydroshare \citep{haacker2023} available at \url{https://www.hydroshare.org/resource/7d925c7944244032af98c9ed20c22db6/}. This data is used for Figure \ref{fig:sat-map} and the regression analysis. Soil characteristics data was obtained from the Soil Survey Geographic Dataset, which is used for regression analysis. The R soilDB package \citep{Rsoildb} was used to download the data. Regression analyses are implemented using the R fixest package \citep{Rfixest}. Other key R packages include ggplot2 \citep{Rggplot2} for creating plots, data.table \citep{Rdatatable} for data wrangling, and sf \citep{Rsf} for spatial data operations. The data used for the regression analysis can be generated using our codes by following the instructions available at the linked Github repository.

\subsection{Code Availability}
All the codes used for this study are available at the following Github repository (\href{https://github.com/tmieno2/Drought-Production-Risk-Aquifer}{link}). All analyses were performed in R \citep{R}.

@@

\subsection*{Competing Interests Statement}

@@

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_yield_with_conf_corn}

@@

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_yield_with_conf_soy}

@@

{\centering \includegraphics[width=6.5in,height=500px,]{../Results/Figures/g_avg_yield}

@@

<<sat_ref, echo = FALSE>>=
sat_ref <-
  readRDS(here::here("Shared/Results/all_results.rds")) %>%
  .[1, sat_seq] %>%
  .[[1]] %>%
  .[length(.)]
@

Figure \ref{fig:test-sat-share} presents the difference in the estimated share of irrigated acres at different levels of aquifer thickness relative to the aquifer thickness level of \Sexpr{sat_ref} m, which represents a high level of aquifer thickness for our study area that exerts no constraints on water supply for irrigation. Grey shaded areas represented the 95\% confidence intervals for the estimated relationships. Where this range exceeds the red zero line, this indicates that the difference in aquifer thickness (relative to \Sexpr{sat_ref} m baseline) leads to a statistically significant (at 5\% level) reduction in irrigated production area share.

@@

{\centering \includegraphics[width=6.5in,height=500px,]{../Results/Figures/g_share_dif}

@@

\caption{Decline in the share of acres that are irrigated as a function of different levels of aquifer thickness relative to a baseline aquifer thickness level of \Sexpr{sat_ref} m (black line). The grey shaded areas represent the 95\% confidence interval for this estimated relationship.}\label{fig:test-sat-share}
\end{figure}

@@

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_variation}

@@

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_share_comp}

@@

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_basis}

@@

{\centering \includegraphics[width=6in,height=500px,]{../Results/Figures/g_il_y}
